\chaplbl{Checklists}{sec:checklists}

See Atul Gawande's 2007 article
``\href{http://www.newyorker.com/magazine/2007/12/10/the-checklist}{The Checklist}''
for a look at how using checklists can save lives (and make many other
things better too).

\seclbl{Scheduling the Event}{sec:scheduling-the-event}

\begin{enumerate}

\item
  Decide if it will be in person, online for one site, or online for
  several.

\item
  Talk through expectations with the host(s).
  If it is in person, make sure the host knows they're covering
  travel costs for trainers.

\item
  Determine who is allowed to attend.

\item
  Arrange trainers.

\item
  Arrange space.
  Make sure there are breakout rooms for video recording.

\item
  Choose dates.
  If it is in person, book travel.

\item
  Get names and email addresses of attendees from host(s).

\item
  Make sure people are added to whatever online registration system is
  being used.

\end{enumerate}

\seclbl{Setting Up}{sec:setting-up}

\begin{enumerate}

\item
  Set up a one-page website for the workshop.

\item
  Add the URL for the workshop website to the tracking system.

\item
  Check whether any attendees have special needs.

\item
  If the workshop is online, test the video conference link.

\item
  Make sure attendees will all have network access.

\item
  Create an Etherpad or Google Doc for shared notes.

\item
  Email attendees a welcome message that includes:

  \begin{itemize}

    \item
      a link to the workshop home page

    \item
      background readings

    \item
      a description of any pre-requisite tasks

  \end{itemize}

\end{enumerate}

\seclbl{During the Event}{sec:during-the-event}

\begin{enumerate}

\item
  Remind everyone of the code of conduct.

\item
  Collect attendance.

\item
  Distribute sticky notes.

\item
  Collect relevant online account IDs (e.g., GitHub IDs).

\item
  Go through the checkout procedure point by point.

\item
  Explain how to contribute to lessons.

\end{enumerate}

\seclbl{After the Event}{sec:after-the-event}

\begin{enumerate}

\item
  Update online records of who participated in what role.

\item
  Administer the post-training survey.

\item
  Email attendees about the checkout process.

\item
  Debrief with the head of instructor training.

\item
  Oversee final demonstrations.

\end{enumerate}

\seclbl{Between Instructor Training Sessions}{sec:between-instructor-training-sessions}

\begin{enumerate}

\item
  Sign up to lead group lesson discussions.

\item
  Monitor lessons for contributions and give feedback.

\end{enumerate}

\seclbl{After Trainees Complete}{sec:after-trainees-complete}

\begin{enumerate}

\item
  Send new instructors a completion message.

\item
  Create certificates and send them to learners.

\end{enumerate}

Note that trainers do not examine their own trainees: having them
examine each other's helps balance load and maintain consistency of
curriculum and standards.
